---
title: "Intermediate Elective 2"
author: Ethan Castelazo
date: 05-13-2026
format: 
  pdf:
    include-in-header: 
       text: |
         \usepackage{fvextra}
         \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
         \DefineVerbatimEnvironment{OutputCode}{Verbatim}{breaklines,commandchars=\\\{\}}
source: editor
---

# Explaining my Inspiration
I chose Steven Ponces' "Coal falls. Renewables rise" as my visualization template because I liked how simplistic and sleek his representaiton of energy data is. This style makes sense for the data I plan to present, because it charts 2 variables (coal vs. renewable energy) that use the same units over the same time in history. I plan to chart % coverage of algae and % coverage of coyote-thistle in the same fashion, to see how the the two variables follow or seperate from each other. From the veg dataset, I'll be using num_quad and percent_cover to calculate a mean_cover for both algae and coyote-thistle, which will be on my y-axis, and I'll use the date column for the x-axis to show change over time.


# 1. Packages

```{r}
#| label: packages
#| warning: false
#| message: false      
#| results: "hide"     

## 1. LOAD PACKAGES & SETUP ----
suppressPackageStartupMessages({
pacman::p_load(
  tidyverse, ggtext, showtext, 
  janitor, scales, glue,
  camcorder, here, magick  # <-- just add magick here
  )
})

```


# 2. Data

```{r}
#| label: reading-data

veg <- read_csv(here("data", "veg.csv"))

metadata <- read_csv(here("data", "vp_veg_metadata.csv"))


```


## Necessary Code from Class

```{r}
#| label: class-wrangling-code

# creating new object from veg
pools_explicit0 <- veg %>% 
  # filtering for vernal pools at ncos sites
  filter(str_detect(transect_name, "VP") &
                    site == "ncos") %>% 
  # handling missing data and filling with 0 values
  complete(year, transect_name, psoc, transect_distance,
           fill = list(percent_cover = 0)) %>% 
  # grouping by year, transect_name, and species
  group_by(year, transect_name, psoc) %>%
  # calculating percent cover per year
  summarize(sum_pc = sum(percent_cover, na.rm = TRUE)) %>%
  #ungrouping the df
  ungroup() %>%
  # joining with metadata to get year_pool
  unite("year_pool", year, transect_name, na.rm = TRUE) %>% 
  left_join(metadata, by = "year_pool") %>% 
  # making a new column of mean percent cover
  mutate(mean_pc = sum_pc/num_quad) %>%
  # grouping by year and species
  group_by(year, psoc) %>%
  # calculating mean percent cover per year across pools
  summarize(mean_pc_year = mean(mean_pc, na.rm = TRUE)) %>% 
  #ungrouping
  ungroup()

# fitering to investigate only coyote-thistle and algae
filtered_explicit0 <- pools_explicit0 %>%
  filter(psoc == "Eryngium_vaseyi" | psoc == "algae")

```


# 3. Tidy Data

```{r}
# retrieving the peak value of algae cover
algae_peak <- filtered_explicit0 %>% 
  filter(psoc == "algae") %>% 
  slice_max(mean_pc_year)

# defining the year when algae peaked
algae_peak_year <- algae_peak$year
# defining the value of peak algae
algae_peak_val  <- round(algae_peak$mean_pc_year, 1)

```



# 4. Visualization Parameters

```{r}
#| label: params
#| include: true
#| warning: false

### |- plot aesthetics ----
colors <- list(palette = list(
  bg_color     = "#0D1B2A",
  coyote_gold   = "#F4A932",
  algae_green    = "#009021",
  text_white   = "#FFFFFF",
  text_muted   = "#8A98A8",
  text_sub     = "#B0B8C4",
  grid_color   = "#1A2E42",
  accent_coral = "#E8503A"
))

bg_color     <- colors$palette$bg_color
coyote_gold  <- colors$palette$coyote_gold
algae_green  <- colors$palette$algae_green
text_white   <- colors$palette$text_white
text_muted   <- colors$palette$text_muted
text_sub     <- colors$palette$text_sub
grid_color   <- colors$palette$grid_color
accent_coral <- colors$palette$accent_coral
col_coyote <- coyote_gold
col_algae  <- algae_green


### |- titles and caption ----
title_text <- glue(
  "<span style='font-weight:900; color:{algae_green};'>Algae Falls.&nbsp;</span>",
  "<span style='font-weight:900; color:{coyote_gold};'> Coyote-thistle persists.</span>"
)

subtitle_text <- glue(
  "<span style='font-size:12pt; font-weight:400; color:{text_sub};'>",
  "Despite mirroring the trajectory of coyote-thistle's percent cover<br>",
  "at  NCOS vernal pools, algae sees a major dropoff in 2024 and 2025",
  "</span>"
)

caption_text <- "Source: Cheadle Center for Biodiversity and Ecological Restoration"

### |- fonts ----
# Load fonts directly — no dependency on his utils/fonts.R
font_add_google("Oswald", "title_font")
font_add_google("Roboto", "text_font")
showtext_auto()
showtext_opts(dpi = 320)

# Create a simple fonts list to match his structure
fonts <- list(
  title = "title_font",
  text  = "text_font"
)
```



# 5. Plot

```{r}
#| label: plot
#| warning: false
#| fig-width: 6
#| fig-height: 10
#| dpi: 320

### |- main plot ----
p <- filtered_explicit0 |>
  ggplot(aes(x = year, y = mean_pc_year, color = psoc, group = psoc)) +
  
  # Geom
  geom_hline(
    yintercept = seq(0, 10, by = 2),
    color = grid_color,
    linewidth = 0.25
  ) +
  geom_line(
    data = filtered_explicit0 |> filter(psoc == "Eryngium_vaseyi"),
    linewidth = 1.9,
    alpha = 1.0
  ) +
  geom_line(
    data = filtered_explicit0 |> filter(psoc == "algae"),
    linewidth = 1.3,
    alpha = 0.60
  ) +
  geom_point(
    data = filtered_explicit0 |> filter(year == 2021),
    size = 3.5,
    shape = 21,
    fill = bg_color,
    stroke = 1.8
  ) +
  geom_point(
    data = filtered_explicit0 |> filter(year == 2025),
    size = 3.5,
    shape = 21,
    fill = bg_color,
    stroke = 1.8
  ) +

    # Scales
  scale_color_manual(
    values = c("Eryngium_vaseyi" = coyote_gold, "algae" = col_algae)
  ) +
  scale_x_continuous(
    limits = c(2021,2025),
    breaks = c(2021, 2022, 2023, 2024, 2025),
    expand = expansion(mult = c(0.03, 0.24))
  ) +
  scale_y_continuous(
    limits = c(0, 10),
    breaks = seq(0, 10, by = 2),
    labels = \(x) glue("{x}%"),
    expand = expansion(mult = c(0.01, 0.04))
  ) +
  
# Annotate
  annotate(
    "segment",
    x = algae_peak_year, xend = algae_peak_year,
    y = 0, yend = algae_peak_val,
    color = accent_coral,
    linewidth = 0.5,
    linetype = "dotted"
  ) +
  annotate(
    "segment",
    x = 2025, xend = 2025,
    y = 4.45, yend = 9.45,
    color = "#D8E4EE",
    linewidth = 0.45,
    linetype = "dotted"
  ) +
  annotate(
    "text",
    x = 2025,
    y = 6.5,
    label = "5% difference\nin pct cover ",
    hjust = 0,
    size = 4,
    color = "#D8E4EE",
    family = fonts$text,
    lineheight = 1.1
  ) +
  annotate(
    "text",
    x = 2021.3,
    y = 1,
    label = glue("Algae"),
    hjust = 0,
    vjust = 0.5,
    size = 3.2,
    color = algae_green,
    family = fonts$text,
    lineheight = 1.15
  ) +
  annotate(
    "text",
    x = 2021.3,
    y = 4,
    label = glue("Coyote-thistle"),
    hjust = 0,
    vjust = 0.5,
    size = 3.2,
    color = coyote_gold,
    family = fonts$text,
    lineheight = 1.15
  ) +
  annotate(
    "text",
    x = 2022.7,
    y = 9,
    label = glue("Algae diverges from\n Coyote-thistle"),
    hjust = 0,
    vjust = 0.5,
    size = 3,
    color = accent_coral,
    family = fonts$text,
    lineheight = 1.15
  ) +



  # Labels
  labs(
    title = title_text,
    subtitle = subtitle_text,
    caption = caption_text,
    x = NULL,
    y = NULL
  ) +

  # Theme
  theme_void() +
  theme(
    # canvas
    plot.background = element_rect(fill = bg_color, color = NA),
    panel.background = element_rect(fill = bg_color, color = NA),
    plot.title = element_markdown(
      family = fonts$title,
      size = 28,
      color = text_white,
      hjust = 0,
      lineheight = 1.1,
      margin = margin(t = 10, b = 6)
    ),
    plot.subtitle = element_markdown(
      family = fonts$text,
      size = 10,
      color = text_sub,
      hjust = 0,
      lineheight = 1.4,
      margin = margin(b = 22)
    ),
    plot.caption = element_markdown(
      family = fonts$text,
      size = 6.5,
      color = "#5A6878",
      hjust = 1,
      margin = margin(t = 14)
    ),
    axis.text.x = element_text(
      family = fonts$text,
      size   = 9,
      color  = text_muted,
      margin = margin(t = 6)
    ),
    axis.text.y = element_text(
      family = fonts$text,
      size = 8,
      color = text_muted,
      hjust = 1
    ),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none",
    plot.margin = margin(t = 20, r = 16, b = 16, l = 20)
  )

p
```

